Unifying intra- and inter-specific variation in tropical tree mortality

### 2017-10-24 ms/ms.Rnw
Updated some of the intext functions and text around subsampling

% Please add here a significance statement to explain the relevance of your work
% 120 words
\significancestatement{Tree mortality is a fundamental demographic process affecting forest dynamics and carbon cycling. Here, for the first time, we use observed survival records over a 15 year period from more than 180,000 individuals, to simultaneously estimate growth-dependent and growth-independent mortality across 203 tropical forest tree species. We found that growth-dependent mortality was the predominate factor influencing tree mortality rates at Barro Colorado Island. Furthermore, we found that wood density most strongly influenced mortality rates by decreasing growth-independent mortality. However, wood density only accounted for a small fraction of the overall species variability observed in growth-independent mortality rates, suggesting that there must be other species traits that strongly affect mortality. Which traits, however, remains unknown.}

@@

% 250 words
\begin{abstract}
Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified factors influencing tree mortality across a variety of scales, these distinct drivers are yet to be integrated within a unified predictive framework. In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive instantaneous mortality functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama. Specifically, we develop mortality functions that not only integrate individual, species and temporal effects, but also simultaneously estimate the independent contribution of growth-dependent and growth-independent hazards on the overall instantaneous mortality rate. We show that functions that partition mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent death, outperform alternative formulations. Moreover, we show that temporal variation in mortality is large and that the effect of wood density -- a prominent trait known to influence tree mortality -- explains only \Sexpr{extract_prop_explained(param_variance_explained,I('wood_density'))}\% of the total variability observed among species. Lastly, our analysis show that growth-dependent processes are the predominant contributor to rates of tree mortality at BCI. Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests. It also highlights how little we know about the causes of species-level and temporal plot-scale effects needed to effectively predict tree mortality.
\end{abstract}

@@

To validate this model, we fit a series of models with increasing complexity (eq. \ref{eq:model} being the most complex) and compare their skill in predicting patterns of tropical tree mortality for \Sexpr{n_inds(BCI_model_dataset_true_dbh)} individuals from \Sexpr{n_spp(BCI_model_dataset_true_dbh)} tree species at Barro Colorado Island (BCI), Panama (Fig. \ref{fig1}). The data are repeat censuses of stem diameter and tree status (alive vs. dead) taken over a 15 yr period (Fig. \ref{fig1}A). We compare the skill of different hazard functions in predicting outcomes in novel data (i.e. not used in model fitting) via 10-fold cross-validation (Fig. \ref{fig1}D) \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations. We therefore fit models across five iterative stages of model development. At each stage, the best model from the previous stage was taken as input for the next stage.

@@

\subsection*{Mortality over time}
Comparing the three 5-year intervals between censuses from 1995 to 2010, we found tree mortality progressively decreasing over time. The highest proportion of trees death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 3)}\%). Consequently, when we allowed hazard rates to be scaled by census (i.e. adding term $\delta$) we observed a significant increase in predictive skill (Fig. \ref{fig2}A). Assuming the long-term average census effect ($\delta$) is 1, deaths were \Sexpr{census_effects(final_model,1)}\% higher in 1995\textendash 2000, and \Sexpr{census_effects(final_model,2)}\% and \Sexpr{census_effects(final_model,3)}\% lower in 2000\textendash 2005 and 2005\textendash 2010, respectively.

@@

Furthermore, future research should focus on examining how climate variables influence both growth-dependent and growth-independent hazards, particularly as climate-driven mortality is increasing globally\cite{vanMantgem:2009bi, Mcdowell:2013wf}. Our analyses revealed a significant decline in mortality at BCI between 1995 and 2010. Unfortunately, we were unable to statistically attribute this temporal change to particular climatic variables due to low temporal replication (N=3). However, the high numbers of deaths that occurred 1995 and 2000, were likely the product of a severe drought caused by an El-Nino in 1997\textendash 1998 \cite{Condit:1995wl, Mcdowell:2013wf}.

@@

Within the 50-ha plot the diameter at breast height and survival status of all free-standing woody plants that were at least 1.3 m tall and had diameter $\geq$ 1 cm were recorded in 1981\textendash1983, 1985, and every 5 years thereafter\cite{Condit:2012nz}. For the purpose of modelling mortality as a function of past growth, we discarded data collected prior to 1990 because diameter measurements were rounded to the nearest 5 mm, whereas later censuses were rounded to the nearest millimetre. Consequently, we modelled tree mortality as a function of past growth for censuses 1995\textendash2000, 2000\textendash2005 and 2005\textendash2010. We discarded species that do not exhibit secondary growth (e.g. palms and ferns), contained fewer than 10 individuals or did not contain an estimate of wood density. We also excluded individuals that: 1) did not survive at least two censuses (two being required to estimate growth rate); 2) were not consistently measured at 1.3 m above ground; 3) were multi-stemmed; 4) resprouted or \textit{``returned from the dead''}; or 5) were extreme outliers -- stems which grew more than 5 cm yr$^{-1}$ or shrunk more than 25\% of their initial diameter. In order to maintain statistical independence of measurements and avoid estimating 180,500 individual parameters, we sub-sampled the dataset by randomly selecting one observation per individual out of a possible three censuses.

>>>>

% Please add here a significance statement to explain the relevance of your work
% 120 words
\significancestatement{Tree mortality is a fundamental demographic process affecting forest dynamics and carbon cycling. Here, for the first time, we use over 400,000 observed survival records collected over a 15 year period from more than 180,000 individuals, to simultaneously estimate growth-dependent and growth-independent mortality across 203 tropical forest tree species. We found that growth-dependent mortality was the predominate factor influencing tree mortality rates at Barro Colorado Island. Furthermore, we found that wood density most strongly influenced mortality rates by decreasing growth-independent mortality. However, wood density only accounted for a small fraction of the overall species variability observed in growth-independent mortality rates, suggesting that there must be other species traits that strongly affect mortality. Which traits, however, remains unknown.}

@@

% 250 words
\begin{abstract}
Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified factors influencing tree mortality across a variety of scales, these distinct drivers are yet to be integrated within a unified predictive framework. In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive instantaneous mortality functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama. Specifically, we develop mortality functions that not only integrate individual, species and temporal effects, but also simultaneously estimate the independent contribution of growth-dependent and growth-independent hazards on the overall instantaneous mortality rate. We show that functions that partition mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent death, outperform alternative formulations. Moreover, we show that the effect of wood density -- a prominent trait known to influence tree mortality -- explains only \Sexpr{extract_prop_explained(param_variance_explained,I('wood_density'))}\% of the total variability observed among species. Lastly, our analysis show that growth-dependent processes are the predominant contributor to rates of tree mortality at BCI. Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests. It also highlights how little we know about the causes of species-level and temporal plot-scale effects needed to effectively predict tree mortality.
\end{abstract}

@@

To validate this model, we fit a series of models with increasing complexity (eq. \ref{eq:model} being the most complex) and compare their skill in predicting patterns of tropical tree mortality for \Sexpr{n_inds(BCI_model_dataset_true_dbh)} individuals from \Sexpr{n_spp(BCI_model_dataset_true_dbh)} tree species at Barro Colorado Island (BCI), Panama (Fig. \ref{fig1}). The data are repeat censuses of stem diameter and tree status (alive vs. dead) taken over a 15 yr period (Fig. \ref{fig1}A). In total \Sexpr{n_obs(BCI_model_dataset_true_dbh)} observations were used to fit these models. We compare the skill of different hazard functions in predicting outcomes in novel data (i.e. not used in model fitting) via 10-fold cross-validation (Fig. \ref{fig1}D) \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations. We therefore fit models across five iterative stages of model development. At each stage, the best model from the previous stage was taken as input for the next stage.

@@

\subsection*{Mortality over time}
Comparing the three 5-year intervals between censuses from 1995 to 2010, we found tree mortality progressively decreasing over time. The highest proportion of trees death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 3)}\%). Consequently, when we allowed hazard rates to be scaled by census (i.e. adding term $\delta$) we observed a small, but significant, increase in predictive skill (Fig. \ref{fig2}A).

@@

Furthermore, future research should focus on examining how climate variables influence both growth-dependent and growth-independent hazards, particularly as climate-driven mortality is increasing globally\cite{vanMantgem:2009bi, Mcdowell:2013wf}. Our analyses revealed a a marginal decline in mortality at BCI between 1995 and 2010. Unfortunately, we were unable to statistically attribute this temporal change to particular climatic variables due to low temporal replication (N=3). However, the high numbers of deaths that occurred 1995 and 2000, were likely the product of a severe drought caused by an El-Nino in 1997\textendash 1998 \cite{Condit:1995wl, Mcdowell:2013wf}.

@@

Within the 50-ha plot the diameter at breast height and survival status of all free-standing woody plants that were at least 1.3 m tall and had diameter $\geq$ 1 cm were recorded in 1981\textendash1983, 1985, and every 5 years thereafter\cite{Condit:2012nz}. For the purpose of modelling mortality as a function of past growth, we discarded data collected prior to 1990 because diameter measurements were rounded to the nearest 5 mm, whereas later censuses were rounded to the nearest millimetre. Consequently, we modelled tree mortality as a function of past growth for censuses 1995\textendash2000, 2000\textendash2005 and 2005\textendash2010. We discarded species that do not exhibit secondary growth (e.g. palms and ferns), contained fewer than 10 individuals or did not contain an estimate of wood density. We also excluded individuals that: 1) did not survive at least two censuses (two being required to estimate growth rate); 2) were not consistently measured at 1.3 m above ground; 3) were multi-stemmed; 4) resprouted or \textit{``returned from the dead''}; or 5) were extreme outliers -- stems which grew more than 5 cm yr$^{-1}$ or shrunk more than 25\% of their initial diameter. In total \Sexpr{n_obs(BCI_model_dataset_true_dbh)} observations were used in this study comprising \Sexpr{n_inds(BCI_model_dataset_true_dbh)} individual trees and \Sexpr{n_spp(BCI_model_dataset_true_dbh)} species. Because of computational difficulties, the models fit in this study do not include individual random effects, as this would require estimation of an additional \Sexpr{n_inds(BCI_model_dataset_true_dbh)} parameters. Instead, our models assume that repeat measurements of an individual are independent of one another. We believe this is a reasonable assumption given that there is approximately 5-years between censuses.
